---
output: github_document
---

```{r, echo = FALSE}
version <- as.vector(read.dcf('DESCRIPTION')[, 'Version'])
version <- gsub('-', '.', version)
```

[![R-CMD-check](https://github.com/tsmodels/tsissm/workflows/R-CMD-check/badge.svg)](https://github.com/tsmodels/tsissm/actions)
[![Last-changedate](https://img.shields.io/badge/last%20change-`r gsub('-', '--', Sys.Date())`-yellowgreen.svg)](/commits/master)
[![packageversion](https://img.shields.io/badge/Package%20version-`r version`-orange.svg?style=flat-square)](commits/master)
[![CRAN_Status_Badge](https://www.r-pkg.org/badges/version/tsissm)](https://cran.r-project.org/package=tsissm)


```{r, echo = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "README-"
)
```

# tsissm

Unobserved components model using the linear innovations state space representation (single source of error). 

Key features:

* Estimation using autodiff (TMB)
* Model selection and ensembling
* Regressors in the observation equation (non-time varying)
* Choice of distributions (Gaussian, Student and Johnson's SU)
* Choice of dynamics in variance (constant or GARCH)

Methods for specification (`modelspec`), estimation (`estimate`), 
summary with choice of vcov (`summary` and `vcov`), diagnostics (`tsdiagnose`), 
online filtering (`tsfilter`), prediction (`predict`), simulation (`simulate`), 
backtesting (`tsbacktest`) and profiling (`tsprofile`). Additionally, model selection (`auto_select`)
and ensembling (`tsensemble`) is also included.
