---
title: "Regular Spacings and Seasonality"
output: 
    rmarkdown::html_vignette:
        css: custom.css
        code_folding: hide
        toc: yes
vignette: >
  %\VignetteIndexEntry{Regular Spacings and Seasonality}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup, warning=FALSE,message=FALSE}
library(tsissm)
library(xts)
library(data.table)
library(tsaux)
library(knitr)
library(zoo)
```

## Introduction

A recent blog [post](https://research.google/blog/autobnn-probabilistic-time-series-forecasting-with-compositional-bayesian-neural-networks/)  
from Google Research introduced a compositional Bayesian neural network model 
and showcased it using the well-known monthly average $CO_2$ concentrations from 
the Mauna Loa observatory. While the example is informative, the monthly series 
poses little challenge from a modeling perspective.

In contrast, this vignette explores the daily Mauna Loa $CO_2$ dataset, which 
introduces substantial complexity due to its irregular sampling — data is not 
available for every day of the year. This irregularity poses problems for seasonal 
modeling, as uneven time intervals distort lag relationships and degrade model inference.

We demonstrate how the `tsissm` package, which natively supports missing values, 
can be used to reframe irregular series onto a regular time grid with gaps — improving 
both seasonal modeling and forecast accuracy.


## Data Investigation

We begin by visualizing the daily Mauna Loa $CO_2$ series:

```{r, fig.width=6,fig.height=3}
data("maunaloa")
opar <- par(mfrow = c(1,1))
par(mar = c(2,2,2,2))
plot(zoo(maunaloa$co2, maunaloa$date), type = "l", xlab = "", ylab = "", main = "Mauna Loa Daily Average CO2")
par(opar)
```

This dataset spans more than 20 years of daily averages. There is a clear seasonal 
pattern and a persistent upward trend. However, a closer look reveals that the 
number of observations per year is not constant:

```{r, fig.width=6,fig.height=3}
d <- data.table(date = maunaloa$date, value = maunaloa$co2)
d[, year := year(date)]
s <- d[,list(.N), by = "year"]
opar <- par(mfrow = c(1,1))
par(mar = c(2,2,2,2))
barplot(s$N, names.arg = s$year, main = "Observations/Year", col = "steelblue")
par(opar)
```

This inconsistency in daily coverage is the root of many issues when trying to fit seasonal models.


## Naive Approach (Irregular Time Base)

A naive approach would be to directly model the series as-is, using a fixed 
seasonal frequency. For this example, we hold out the years 2024–2025 for evaluation:

```{r, fig.width=6,fig.height=3}
co2 <- xts(maunaloa$co2, maunaloa$date)
train <- co2["/2023"]
test <- co2["2024/"]
spec_naive <- issm_modelspec(train, slope = TRUE, seasonal = TRUE, seasonal_frequency = 320, seasonal_harmonics = 2)
# fixed slope
spec_naive$parmatrix[parameters == "beta", initial := 0]
spec_naive$parmatrix[parameters == "beta", lower := 0]
spec_naive$parmatrix[parameters == "beta", estimate := 0]
mod_naive <- estimate(spec_naive, scores = FALSE)
p_naive <- predict(mod_naive, h = length(test), nsim = 3000, forc_dates = index(test))
opar <- par(mfrow = c(1,1))
par(mar = c(2,2,2,2))
plot(p_naive, n_original = 600, gradient_color = "yellow", interval_color = "orange", main = "CO2 [Naive] Model Prediction", median_width = 1, xlab = "")
lines(index(test), as.numeric(test), col = 3)
legend("topleft", c("Median Forecast","Actual"), col = c(1,3), lty = 1, bty = "n")
par(opar)
```

The resulting forecast exhibits a phase shift in the seasonal component. 
This is expected — irregular time spacing distorts seasonality, resulting in 
inaccurate timing of peaks and troughs.

## Correct Approach (Regular Time Base with Missing Values)

A better solution is to recast the series on a regular daily time grid, filling 
in missing observations with NA. Since `tsissm` supports missing values in the 
response variable $y_t$, this allows us to model the series in calendar time 
while preserving seasonal structure.

In `tsissm`, if y_t is missing, the innovation $\varepsilon_t$ is set to 
zero — its expected value — effectively treating the model output as a 
one-step-ahead forecast. These missing values are ignored in the likelihood and 
handled correctly during state filtering.

```{r}
full_dates <- seq(index(co2)[1], tail(index(co2),1), by = "days")
co2_full <- xts(rep(as.numeric(NA), length(full_dates)), full_dates)
co2_full <- cbind(co2_full, co2, join = "left")
co2_full <- co2_full[,-1]
```

This new dataset now spans a regular daily grid. The percentage increase in data 
points (mostly NA) is: `r print(paste0(round(100 * (NROW(co2_full)/NROW(co2) - 1),0), "%"))` more data points which are all missing values.

## Re-estimating with Regularized Calendar Time

```{r, fig.width=6,fig.height=3}
train_full <- co2_full["/2023"]
test_full <- co2_full["2024/"]
spec <- issm_modelspec(train_full, slope = TRUE, seasonal = TRUE, seasonal_frequency = 366, seasonal_harmonics = 5, distribution = "norm")
# fixed slope
spec$parmatrix[parameters == "beta", initial := 0]
spec$parmatrix[parameters == "beta", lower := 0]
spec$parmatrix[parameters == "beta", estimate := 0]
mod <- estimate(spec, scores = FALSE)
p <- predict(mod, h = length(test_full), nsim = 3000, forc_dates = index(test_full))
opar <- par(mfrow = c(1,1))
par(mar = c(2,2,2,2))
plot(p, n_original = 600, gradient_color = "yellow", interval_color = "orange", main = "CO2 [Correct] Model Prediction", median_width = 1, xlab = "")
lines(index(test_full), as.numeric(test_full), col = 3)
legend("topleft", c("Median Forecast","Actual"), col = c(1,3), lty = 1, bty = "n")
par(opar)
```

The forecast now clearly aligns better with the actual seasonal pattern — the 
phase error seen earlier has been corrected.

## Evaluation

We compare forecast accuracy between the naive and corrected models using MAPE and CRPS. 
Only non-missing values are considered:

```{r}
use <- which(!is.na(as.numeric(test_full)))
tab <- data.frame(MAPE = c(mape(as.numeric(test), as.numeric(p_naive$analytic_mean)) * 100, 
                           mape(as.numeric(test_full)[use], as.numeric(p$analytic_mean)[use]) * 100),
                  CRPS = c(crps(as.numeric(test), p_naive$distribution), crps(as.numeric(test_full)[use], p$distribution[,use])))
rownames(tab) <- c("Naive","Correct")
colnames(tab) <- c("MAPE (%)", "CRPS")
tab |> kable(digits = 2)
```

The table shows a clear improvement in both metrics for the correctly specified model — particularly a halving of the MAPE.

## Conclusion

Irregular sampling in time series — especially those with seasonal structure — can 
significantly impair model inference and forecasting. This vignette demonstrated 
how naive models fitted on irregular time bases can introduce phase errors and bias. 
By recasting the series onto a regular grid with missing values, and leveraging 
`tsissm’s` built-in support for such data, we were able to restore seasonal 
fidelity and greatly improve forecast accuracy.

This approach is general and applicable to many real-world time series where 
regular sampling cannot be guaranteed.

